---
title: "Data Wrangling in tidyr"
author: "Amritpal Sandhu"
date: "2022-08-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r}
library(tidyverse)
library(here)
library(janitor)
```

## Read in World Bank (wb) data files

```{r}
wb_indicators <- read_csv(here::here("data", "wb_indicators.csv"), na = c("..", ""))

# Make the ".." into na's up above and if there is nothing there it should also be na. 

wb_metadata <- read_csv(here::here("data", "wb_indicators_metadata.csv"))


```

## Tidy up my data

### pivot_longer to get years into a single column

```{r}
# Look up names(wb_indicators) in the console to get the exact names for the columns

wb_indicators_long <- wb_indicators |> 
  pivot_longer(cols = '2001 [YR2001]' : '2020 [YR2020]',
               names_to = "year", 
               values_to = "indicator_value")

# You can also check class of year by: class(wb_indicators_long$year)

```

### Separate to clean up the year column

#### You can also pipe through the whole this if you wanted, but we are breaking it up for now

```{r}
wb_clean <- wb_indicators_long |>
  separate(col = year, into = c("year", "year_chr"), sep = " ") |> 
  select(-year_chr, -'Country Code', -'Series Code') |>
  mutate(year = as.numeric(year)) |>
  drop_na('Series Name') |> 
  pivot_wider(names_from = 'Series Name', values_from = indicator_value)

# The select function helped remove the columns we don't want and then mutate changed the year values to numeric. drop_na means any row with na can be dropped and this is used for rows. 
  
# Next we widen the variables currently trapped in the 'Series Name" to be spread across multiple columns, and get the values that populated the columns from 'indicator value'


# Rename the columns 

names(wb_clean) <- c("country", "year", "acess_clean_fuels_pp", "access_electricity_pp", "co2_emissions_kt", "fossil_fuel_cons_pct", "water_stress")

wb_clean_2 <- wb_clean |> rename(my_year = year)



wb_subset <- wb_clean |>
  filter(country %in% c("Algeria", "Barbados", "Bulgaria", "Chile"))

# Then you can check unique in the console to check if you filtered correctly: unique(wb_subset)

# Make a ggplot

ggplot(data = wb_subset, aes(x = year, 
                             y = co2_emissions_kt, group = country)) + geom_line() + facet_wrap(~country)

# You can also group through geom_line(aes(color = country))
```

```{r}
# The ! means to exclude 


wb_clean_2 <- wb_indicators_long |>
  separate(col = 'year', into = c("year", "year_chr"), sep = " ")
```

## Filtering Rows

```{r}
# Example 1: Filter to keep exact matches

# Height greater than 180 
ex_1 <- starwars |>
  filter(height > 180)

min(ex_1$height)

```

```{r}
## Use double equal to get exact value

ex_2 <- starwars |>
  filter(eye_color == "blue")

unique(ex_2$eye_color)
```

```{r}
# Make a subset that only contains rows where the homeworld is Naboo

ex_3 <- starwars |>
  filter(homeworld == "Naboo")
```

## Examples: filtering to satisfy multiple examples

```{r}
# Height greater than 180 and homeworld is tatooine

ex_4 <- starwars |>
  filter(height > 180 & homeworld == "Tatooine")

```

```{r}
# Create a subset where hair color brown and species is human 

ex_5 <- starwars |>
  filter(hair_color == "brown" & species == "Human")
```

```{r}
# Filtering OR statements

ex_6 <- starwars |>
  filter(height > 180 | eye_color == "yellow")
```


```{r}
# Create a subset that keeps rows if homeworld is Endor or species is droid

ex_7 <- starwars |>
  filter(homeworld == "Endor" | species == "Droid")
```


```{r}
ex_8 <- starwars |>
  filter(eye_color == "blue" | eye_color == "brown" | eye_color == "red")

## Another way to do this is through in operator

ex_9 <- starwars |>
  filter(eye_color %in% c("blue", "brown", "red"))

# NEVER do this because double equal only asks to match values in the order that you have written. Only use if looking for a single exact match  

ex_10 <- starwars |>
  filter(eye_color == c("blue", "brown", "red"))
```

```{r}
# Create a subset where homeworld is Endor OR Tatooine OR Naboo AND height < 150

ex_11 <- starwars |>
  filter(homeworld %in% c("Endor", "Tatooine", "Naboo"),
         height < 150)

```

```{r}
# All species except human
ex_12 <- starwars |>
  filter(species != "Human")

# All species except human, ewok, and wookie

ex_13 <- starwars |> 
  filter(!species %in% c("Human", "Ewok", "Wookie"))
```

## `select()` : select or exclude columns

```{r}
# Select by name

ex_14 <- wb_clean |>
  select(country, year, co2_emissions_kt)

ex_15 <- wb_clean |>
  select(year:fossil_fuel_cons_pct)

# Exclude the electricity 

ex_16 <- wb_clean |>
  select(year:fossil_fuel_cons_pct, -access_electricity_pp)
```

```{r}
# Filter the countries and all the columns from country to fossil fuel

ex_17 <- wb_clean |>
  filter(country %in% c("Bolivia", "Chile", "Mexico")) |>
  select(country:fossil_fuel_cons_pct)
```

```{r}
# Change the column names

ex_18 <- wb_clean |>
  select(year_new = year,
         emissions = co2_emissions_kt, 
         country_name = country)

# Select only columns from country, year, and water stress, and reorder them as year, country, water stress, then filter to only include observations for three countries of your choosing. 

ex_19 <- wb_clean |>
  select(year, country, water_stress) |>
  filter(country %in% c("Australia", "Aruba", "Lebanon"))

# Check the filter for country
unique(ex_19$country)
```

```{r}
# only renaming 2 of the columns. Remember new name = old name

ex_20 <- wb_clean |>
  rename(ffcp = fossil_fuel_cons_pct,
         ws = water_stress)
```

```{r}
# rename 2 columns

ex_21 <- wb_clean |>
  rename(co2 = co2_emissions_kt, 
         the_country = country)
```

### `mutate()` : update an existing column, or add a new one

If you assign a NEW column name, it will add a new column to the end of your data frame. If you assign an existing column name, it will overwrite that column. 

```{r}
# Create new column, add another column and include the calculation 

ex_22 <- wb_clean |>
  mutate(co2_emissions_t = co2_emissions_kt * 1000) |> relocate(co2_emissions_t, .after = co2_emissions_kt)

ex_23 <- wb_clean |>
  mutate(yay = "YAY")

# Changed class to character 

ex_24 <- wb_clean |>
  mutate(year = as.character(year))

# Change back to numeric

ex_25 <- ex_24 |>
  mutate(year = as.numeric(year))

# Use dplyr::relocate()
# Moving water stress next to the year column in the table
ex_26 <- wb_clean |>
  relocate(water_stress, .after = year)

```

### `group_by |> summarize`

group_by: Creates groups under the hood based in variables you 

summarize: apply your functions to EACH GROUP, and return them into a nice table

```{r}
# We are just looking at species and looking for the mean height and mass

ex_27 <- starwars |>
  group_by(species) |>
  summarize(mean_height = mean(height, na.rm = TRUE),
            mean_mass = mean(mass, na.rm = TRUE))

# We want to add the filter through home

ex_28 <- starwars |>
  filter(homeworld %in% c("Naboo", "Tatooine")) |>
  group_by(species) |>
  summarize(mean_height = mean(height, na.rm = TRUE),
            mean_mass = mean(mass, na.rm = TRUE))
```

```{r}
# Make a data frame

df <- tribble(
  ~species, ~length, ~number,
  "lobster", 12, 4, 
  "lobster", 14, 6, 
  "lobster", 15, 2
)

# Make sure every observation has it's own row

df_case <- df |>
  uncount(number)
```

```{r}
# count counts obervations
# combination of group_by summarize n()

starwars |>
  count(homeworld)

starwars |>
  count(homeworld, species)

starwars |>
  group_by(homeworld, species)|>
  summarize (size = n())
```

### Practice example:

- Start from wb_clean
Only keep observations for Nicaragua, Ecuador, Peru, and Chile, AND for years greater than 2005
- Only keep columns for every country, year and co2 emissions
- Add a new column that is co2 emissions converted to megatons
- Find the total co2 emissions in megatons for years still in the dataset by country 

```{r}
ex_29 <- wb_clean |>
  filter(country %in% c("Nicaragua", "Ecuador", "Peru", "Chile"), year > 2005) |> 
  select(country, year, co2_emissions_kt) |>
  mutate(co2_emissions_mt = co2_emissions_kt / 1000) |>
  group_by(country) |>
  summarize(total_co2_mt = sum(co2_emissions_mt), na.rm = TRUE)

unique(ex_29$country)
min(ex_29$year)
```

